Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-07-30 main.Rmd
polishing the MMPD section

\noindent The data structure suggests that we look at pairs of cyclic granularities at a time from $N_C$ cyclic granularities. Hence, all possible pairs of cyclic time granularities that could be analyzed together is $^{N_C} P_2$, with each of the circular granularity either mapped to facets or x-axis. Now, if $N_C$ is fairly large, even after excluding clashes, the list of harmonies left would be very large and overwhelming for human consumption. Hence, there is a need to rank the harmonies basis how well they capture the variation in the measured variable and additionally reduce the number of harmonies for further exploration/visualization. Displays that capture more variation within different categories in the same group would be important to bring out different patterns of the data. Hence, the idea  would be to select only those harmonies which have significant variation in the measured variable and to rank the selected harmonies, such that a harmony pair is rated more important if the variation between different levels of the x-axis variable is higher on an average across all levels of facet variables. This twofold objective is achieved by defining a statistic that essentially captures the variation in a harmony pair and a threshold below which the variation is not considered significant.

@@

\noindent One of the potential ways to evaluate this variation is by computing the maximum pairwise distances between the distributions of the measured variable. <!--We consider distances between probability distributions which lines up with our pivotal idea of being able to visualize probability distributions across pairs of cyclic granularities.-->
One of the common ways to measure distances between distributions is through Jenson-Shanon divergence (@Lin1991-fj) which is based on Kullback-Leibler divergence (@Kullback1951-jy). The Jensen-Shanon distance between two probability distribution $p_1$ and $p_2$ is given by $$d = [D(p_1, r) + D(p_2, r)]/2 \quad where \quad r = (p_1 + p_2)/2$$ where,
 $$D(p_1,p_2) = \int^{\infty}_{-\infty}p_1(x)log\frac{p_1(x)}{p_2(x)}\,dx$$ is the Kullback-Leibler divergence between $p_1$ and $p_2$. Probability distributions are estimated through quantiles instead of kernel density so that there is minimal dependency on selecting kernel or bandwidth.

@@

The tsibble object `smart_meter10` from R package `gravitas` [@R-gravitas] is used to facilitate the systematic exploration here. While trying to explore the energy behavior of these customers systematically across cyclic time granularities, the first thing to consider is which cyclic time granularities we can look at exhaustively. Let us consider conventional time deconstructions for a Gregorian calendar (second, minute, half-hour, hour, day, week, month, year). Since the interval of this tsibble is 30 minutes, the temporal granularities may range from half-hour to year. Considering $6$ linear granularities half-hour, hour, day, week, month and year in the hierarchy table, $N_C = (6*5/2) = 15$. <!--(via Section \ref{sec:data-structure}) circular or quasi-circular granularities that could be formed relating two linear granularities at a time.--> If $N_C$ seem too large, the smallest and largest linear granularities could be considered to be removed from the hierarchy table. We remove half-year and year to have $N_C = (4*3/2) = 6$ and obtain cyclic granularities namely "hour_day", "hour_week", "hour_month", "day_week", "day_month" and "week_month", read as "hour of the day", etc. Further, we add cyclic granularity day-type( "wknd_wday") to capture weekend and weekday behavior. Now that we have a list of cyclic granularities to look at, we should be able to compute the multiple-order-up granularities using Section \ref{sec:cyclic-calendar}. <!-- The lubridate package [@Grolemund2011-vm] provides tools to compute single-order-up granularities, \ref{sec:circular-gran-def} (for hour_day, hour_week and day_week) and \ref{sec:quasi-circular-gran-def} (for hour_month, day_month, week_month).  
with the index set mapping half-hourly time periods to a set of positive integers.-->

@@

From the search list, $N_C = 7$ cyclic granularities are chosen for which we would like to derive insights of energy behavior. Recalling the data structure <$C_i$, $C_j$, `general_supply_kwh`> for exploration $\forall i, j \in \{1, 2, \ldots, 7\}$, each of these $7$ cyclic granularities can either be mapped to x-axis or to facet. Choosing $2$ of the possible $7$ granularities, which is equivalent to having $^{7}P_2 = 42$ candidates for visualization. Fortunately, harmonies can be identified among those $42$ possibilities to narrow the search. \autoref{tab:harmony-tab} shows $10$ harmony pairs after removing clashes and any cyclic granularities with levels more than 31, as effective  exploration becomes difficult with many levels (Section \ref{sec:levels}). The MMPD is also shown along with a threshold indicator taking the value TRUE only when variation is significant. Starting from $42$ possible pairs of cyclic granularities to visualize, we are finally left with only $6$, which is a very sizable number of displays for exploration.

Few harmony pairs are displayed in \autoref{fig:bothcust} to illustrate the siginificance of MMPD, threshold and the impact of different distribution plots or reverse mapping.
\noindent For each of \autoref{fig:bothcust} (b) and (c), $C_i$ is the circular granularity day-type (weekday/weekend) and $C_j$ is hour of the day. The geometry used for displaying the distribution is chosen as area-quantiles and violins in \autoref{fig:bothcust} (b and c respectively). \autoref{fig:bothcust} (a) displays reverse mapping of $C_i$ and $C_j$ with $C_i$ denoting hour of the day and $C_j$ denoting day-type with distribution geometrically displayed as boxplots.

@@

  harmonies <- sm %>%
  harmony(ugran = "month",
          filter_in = "wknd_wday", 
          filter_out = c("hhour", "fortnight"))

global_harmony <- sm %>%
  global_threshold(harmony_tbl = harmonies,
                   response = "general_supply_kwh")%>%
  # select(facet_variable,x_variable,
  #        facet_levels,x_levels,MMPD) %>% 
  rename(`facet variable` = facet_variable,
         `x-axis variable` = x_variable,
         `facet levels` = facet_levels,
         `x-axis levels` = x_levels,
         `average maximum variation` = MMPD)

@@

  knitr::kable(smart_harmony,
               format = "latex",
               booktabs = TRUE,
                caption = "Harmonies with two cyclic granularity one placed on facet and the other on x-axis along with their number of categories/levels. Out of 42 possible combinations of cyclic granularities, only these 10 are harmony pairs. The harmony pairs are ranked basis highest to lowest MMPD, only the first 6 of which are significant.") %>%
    #row_spec(0, bold = TRUE) %>%
    kable_styling()
```

@@

In \autoref{fig:bothcust} (b), the distribution of energy consumption is plotted across the harmony pair (weekday/weekend, hour of the day) through an area quantile plot. The black line is the median, whereas the purple band covers 25th to 75th percentile, the orange band covers 10th to 90th percentile and the green band covers 1st to 99th percentile. The first facet represents the weekday behavior while the second one displays the weekend behavior and energy consumption across each hours of the day is shown inside each facet. The energy consumption is extremely (positive- or right-) skewed with the 1st, 10th and 25th percentile lying relatively close whereas 75th, 90th and 99th lying further away from each other. This is common across both weekdays and weekends. For the first few hours on weekdays, median energy consumption starts and continues to be higher for longer as compared to weekends.

@@

This case study shows systematic exploration of energy behavior for a random household to gain some insights on periodic behavior of the households. 
<!--First, it helps us to find the list of cyclic granularities to look at, then shrinks the number of possible visualizations by identifying harmonies, visualize a harmony pair and shows the effect of different distribution plots or reverse mapping.-->

@@

```{r cricex, fig.cap= "Runs per over shown with different distribution displays, and granularities. Plot (a) shows letter value plot across overs faceted by innings. For the team batting in the first inning there is an upward trend of runs per over, while there is no clear upward trend in median and quartile deviation of runs for the teams batting in the second inning. Plot (b) shows box plot of runs per over across an indicator of wickets in previous over faceted by current over. This indicates that at least one wicket in the previous over leads to lower median run rate and quartile spread in the subsequent over.", warning = FALSE, message = FALSE, out.width = "90%"}

>>>>

\noindent The data structure suggests that we look at pairs of cyclic granularities at a time from $N_C$ cyclic granularities. Hence, all possible pairs of cyclic time granularities that could be analyzed together is $^{N_C} P_2$, with each of the circular granularity either mapped to facets or x-axis. Now, if $N_C$ is fairly large, even after excluding clashes, the list of harmonies left would be very large and overwhelming for human consumption. Hence, there is a need to rank the harmonies basis how well they capture the variation in the measured variable and additionally reduce the number of harmonies for further exploration/visualization. Displays that capture more variation within different categories in the same group would be important to bring out different patterns of the data. Hence, the idea  would be to select only those harmonies which have significant variation in the measured variable and to rank the selected harmonies, such that a harmony pair is rated more important if the variation between different levels of the x-axis variable is higher on an average across all levels of facet variables. This twofold objective is achieved by defining a statistic that could essentially capture the variation of the measured variable in a harmony pair and constructing a test to check if the variation is statistically significant.

@@

\noindent One of the potential ways to evaluate this variation is by computing the maximum pairwise distances between the distributions of the measured variable. <!--We consider distances between probability distributions which lines up with our pivotal idea of being able to visualize probability distributions across pairs of cyclic granularities.-->
A common way to measure distances between distributions is through Jenson-Shanon divergence (@Lin1991-fj) which is based on Kullback-Leibler divergence (@Kullback1951-jy). The Jensen-Shanon distance between two probability distribution $p_1$ and $p_2$ is given by $$d = [D(p_1, r) + D(p_2, r)]/2 \quad where \quad r = (p_1 + p_2)/2$$ where,
 $$D(p_1,p_2) = \int^{\infty}_{-\infty}p_1(x)log\frac{p_1(x)}{p_2(x)}\,dx$$ is the Kullback-Leibler divergence between $p_1$ and $p_2$. Probability distributions are estimated through quantiles instead of kernel density so that there is minimal dependency on selecting kernel or bandwidth.

@@

The tsibble object `smart_meter10` from R package `gravitas` [@R-gravitas] consisting of `reading_datetime`, `customer_id` and `general_supply_kwh` denoting the index, key and measured variable of the tsibble is used to facilitate the systematic exploration. While trying to explore the energy behavior of these customers systematically across cyclic time granularities, the first thing to consider is which cyclic time granularities we can look at exhaustively. Let us consider conventional time deconstructions for a Gregorian calendar (second, minute, half-hour, hour, day, week, month, year). Since the interval of this tsibble is 30 minutes, the temporal granularities may range from half-hour to year. Considering $6$ linear granularities half-hour, hour, day, week, month and year in the hierarchy table, $N_C = (6*5/2) = 15$. <!--(via Section \ref{sec:data-structure}) circular or quasi-circular granularities that could be formed relating two linear granularities at a time.--> If $N_C$ seem too large, the smallest and largest linear granularities could be considered to be removed from the hierarchy table. We remove half-year and year to have $N_C = (4*3/2) = 6$ and obtain cyclic granularities namely "hour_day", "hour_week", "hour_month", "day_week", "day_month" and "week_month", read as "hour of the day", etc. Further, we add cyclic granularity day-type( "wknd_wday") to capture weekend and weekday behavior. Now that we have a list of cyclic granularities to look at, we should be able to compute the multiple-order-up granularities using Section \ref{sec:cyclic-calendar}. <!-- The lubridate package [@Grolemund2011-vm] provides tools to compute single-order-up granularities, \ref{sec:circular-gran-def} (for hour_day, hour_week and day_week) and \ref{sec:quasi-circular-gran-def} (for hour_month, day_month, week_month).  
with the index set mapping half-hourly time periods to a set of positive integers.-->

@@

From the search list, $N_C = 7$ cyclic granularities are chosen for which we would like to derive insights of energy behavior. Recalling the data structure <$C_i$, $C_j$, `general_supply_kwh`> for exploration $\forall i, j \in \{1, 2, \ldots, 7\}$, each of these $7$ cyclic granularities can either be mapped to x-axis or to facet. Choosing $2$ of the possible $7$ granularities, which is equivalent to having $^{7}P_2 = 42$ candidates for visualization. Fortunately, harmonies can be identified among those $42$ possibilities to narrow the search. \autoref{tab:harmony-tab} shows $16$ harmony pairs after removing clashes and any cyclic granularities with levels more than $31$, as effective  exploration becomes difficult with many levels (Section \ref{sec:levels}). The MMPD is also shown along with indicator (*) only when variation of measured variable across the harmony pair significant. Starting from $42$ possible pairs of cyclic granularities to visualize, we are finally left with only $6$, which is a very sizable number of displays for exploration.

Few harmony pairs are displayed in \autoref{fig:bothcust} to illustrate the siginificance of MMPD, threshold and the impact of different distribution plots and reverse mapping.
\noindent For each of \autoref{fig:bothcust} (b) and (c), $C_i$ is the circular granularity day-type (weekday/weekend) and $C_j$ is hour of the day. The geometry used for displaying the distribution is chosen as area-quantiles and violins in \autoref{fig:bothcust} (b and c respectively). \autoref{fig:bothcust} (a) displays reverse mapping of $C_i$ and $C_j$ with $C_i$ denoting hour of the day and $C_j$ denoting day-type with distribution geometrically displayed as boxplots.

@@

  harmonies <- sm %>%
  harmony(ugran = "month",
          filter_in = "wknd_wday", 
          filter_out = c("hhour", "fortnight"))
set.seed(12345)

global_harmony <- sm %>%
  global_threshold(harmony_tbl = harmonies,
                   response = "general_supply_kwh", nsamp = 5) %>%
  rename(`facet variable` = facet_variable,
         `x-axis variable` = x_variable,
         `facet levels` = facet_levels,
         `x-axis levels` = x_levels) %>% 
  mutate(MMPD = if_else(gt_MMPD == TRUE,
                        paste0(MMPD, "*"), paste0(MMPD))) %>%
  select(`facet variable`,`x-axis variable`,
         `facet levels`,`x-axis levels` ,MMPD)

@@

  knitr::kable(global_harmony,
               format = "latex",
               booktabs = TRUE,
                caption = "Harmonies with a pair of cyclic granularity one placed on facet and the other on x-axis. Out of 42 possible combinations of cyclic granularities, only 16 are harmony pairs. The harmony pairs are ranked basis highest to lowest MMPD and only the first 6 show significant variation in the measured variable.") %>%
    #row_spec(0, bold = TRUE) %>%
    kable_styling()
```

@@

In \autoref{fig:bothcust} (b), <!--the distribution of energy consumption is plotted across the harmony pair (weekday/weekend, hour of the day) through an area quantile plot-->the black line is the median, whereas the purple band covers 25th to 75th percentile, the orange band covers 10th to 90th percentile and the green band covers 1st to 99th percentile. The first facet represents the weekday behavior while the second one displays the weekend behavior and energy consumption across each hours of the day is shown inside each facet. The energy consumption is extremely (positive- or right-) skewed with the 1st, 10th and 25th percentile lying relatively close whereas 75th, 90th and 99th lying further away from each other. This is common across both weekdays and weekends. For the first few hours on weekdays, median energy consumption starts and continues to be higher for longer as compared to weekends.

@@

This case study shows systematic exploration of energy behavior for a household to gain exhaustive insights on periodic behavior of the households. 
<!--First, it helps us to find the list of cyclic granularities to look at, then shrinks the number of possible visualizations by identifying harmonies, visualize a harmony pair and shows the effect of different distribution plots or reverse mapping.-->

@@

```{r cricex, fig.cap= "Runs per over shown with different distribution displays, and granularities. Plot (a) shows letter value plot across overs faceted by innings. For the team batting in the first innings there is an upward trend of runs per over, while there is no such pattern of runs for the teams batting in the second innings. Plot (b) shows quantile plot of runs per over across an indicator of wickets in previous over faceted by current over. This indicates that at least one wicket in the previous over leads to lower median run rate and quartile spread in the subsequent over.", warning = FALSE, message = FALSE, out.width = "90%"}
